Semantic Linearity in a Lazy Optimising Compiler

### 2025-01-24 icfp.tex
Move new section to live .tex file

>>>>

\section{Instrumented Operational Semantics}

In this section we develop a lazy big-step operational semantics for $\lambda_\Delta^\pi$ in the
style of Launchbury~\cite{10.1145/158511.158618}. The semantics is
equipped with an environment which assigns values to variables and is
mutated in order to express lazy evaluation. Terms are explicit about
their sharing (in the sense of shared reductions in lazy evaluation),
and so we perform a translation step that makes all sharing explicit
through let-binders.
Below we write $\llet{x{:}_{\Delta}\sigma = e,\dots,y{:}_{\Delta'}=e''
}{e'}$ for the iterated let-binding of variables $x , \dots , y$.       
\[
  \begin{array}{lclr}
    x^* & \triangleq & x\\

    (\Lambda p.~e)^* & \triangleq & \Lambda p.~(e)^*\\
    (\lambda x:_\pi\sigma .e)^* & \triangleq & \lambda x:_\pi\sigma . (e)^*\\
    (e~\pi)^* & \triangleq & (e)^*~\pi\\
    (e~x)^* & \triangleq & (e)^*~x\\
    (e~e')^* & \triangleq & \llet{x{:}_{\Delta}\sigma =
                            (e')^*}{(e)^*~x}     \\ &&\qquad \mbox{with
                            $\Gamma;\Delta' \vdash e :
                            \sigma\rightarrow_\pi \varphi$ and
                                                   $\Gamma;\Delta \vdash e' : \sigma$}\\
    K~e_1~\dots~e_n & \triangleq & \llet{x_1:_{\Delta_1} \sigma_1 =
                                   (e_1)^*,\dots,x_n:_{\Delta_n}
                                   \sigma_n =
                                   (e_n)^*}{K~x_1~\dots~x_n} \\ &&
                                                               \qquad\mbox{with
                                                               $\Gamma;\Delta_1\vdash
                                                               e_1 :
                                                               \sigma_1
                                                               \dots \Gamma;\Delta_n\vdash
                                                               e_n :
                                                               \sigma_n$}\\                                              
(\ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\rho\to e'}\}})^* &
                                                               \triangleq
                                &
                                  \ccase{(e)^*}{z{:}_{\Delta}\sigma~\{\overline{\rho\to (e')^*}\}}\\
    (\llet{x{:}_{\Delta}\sigma = e}{e'}      )^* & \triangleq & \llet{x{:}_{\Delta}\sigma  = (e)^*}{(e')^*}    \\
    (\lletrec{\overline{x{:}_{\Delta}\sigma = e}}{e'}  )^* & \triangleq &\lletrec{\overline{x{:}_{\Delta}\sigma = (e)^*}}{(e')^*}  
      \end{array}
\]

The big-step operational semantics is defined via the relation
$\Gamma : e \Downarrow \Delta : v$ where $e$ is an expression, $v$ a
value, $\Gamma$ and $\Delta$ are environments with bindings of the
form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
variable $x$ of the given type.

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma : \Lambda p.e \Downarrow \Gamma : \Lambda p.e}
    \qquad
    \infer[]
    {\Gamma : e\Downarrow \Delta : \Lambda p.e' \quad \Delta :
    e'[\pi/p]\Downarrow \Theta : v}
    {\Gamma : e~\pi\Downarrow \Theta : v }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma : \lambda x:_\pi \sigma . e \Downarrow \Gamma : \lambda x:_\pi \sigma . e }
    \qquad
    \infer[]
    {\Gamma : e \Downarrow \Delta : \lambda y:_\pi \sigma . e' \quad
    \Delta : e'[x/y] \Downarrow \Theta : v }
    {\Gamma : e~x \Downarrow \Theta : v}\\[1.5em]
    \infer[]
    {\Gamma : e \Downarrow \Delta : v}
    {(\Gamma , x:_\omega \sigma = e) : x \Downarrow (\Delta,x:_\omega
    \sigma = v) : v}
    \qquad
    \infer[]
    {(\Gamma,x:_\omega \sigma = e) : e' \Downarrow \Theta : v}
    {\Gamma : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta : v
    }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma : K~x_1~\dots~x_n \Downarrow \Gamma : K~x_1~\dots~x_n}
    \qquad
    \infer[]
    {\Gamma : e \Downarrow \Delta : K~x_1~\dots~x_n \quad
    \Delta : e'[x_i/y_i][K~x_1~\dots~x_n/z] \Downarrow \Theta : v }
    {\Gamma :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}}
    \Downarrow \Theta : v}\\[1.5em]
    \infer[]
    {\Gamma : e \Downarrow \Delta : K~x_1~\dots~x_n \quad
    \Delta : e'[K~x_1~\dots~x_n/z] \Downarrow \Theta : v }
    {\Gamma :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\} \Downarrow \Theta : v}
    \qquad
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta
    : v
    }
    {\Gamma : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
    \Downarrow \Theta : v}
    \end{array}
\]

We follow an approach
similar to that of \cite{cite:linearhaskell}, defining a linearity
aware operational semantics in which linear variables are erased from
the runtime environment, ensuring that any term violating linearity
will result in a stuck state. The semantics is type-aware, defined by
the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma$ where $\Gamma$ and $\Delta$ are typing judgments
($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
$x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$.

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
    \Lambda p.e) : \forall p . \sigma }
    \\[1.5em]
    \infer[]
    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
    \Lambda p.e') : \forall p . \sigma \quad \Gamma ;\Delta \vdash
    (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' : v) : \sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
    \mid v) : \sigma  }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
    \rightarrow_\pi \tau }\\[1.5em]
    \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_\pi \sigma . e') : \sigma \rightarrow_\pi \tau
    \quad
    \Gamma ; \Delta \vdash (\Theta' \mid e'[x/y]) \Downarrow (\Theta''
    : v) : \tau }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta'
    \mid v) : \tau}\\[1.5em]

       \infer[]
    {\Gamma , x:_{\Delta'} \sigma ; \Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
    \Downarrow (\Theta' , x:_{\Delta'} \sigma = v \mid v) : \sigma}
    \\[1.5em]
           \infer[]
    {\Gamma ;\Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
    \Downarrow (\Theta' \mid v) : \sigma}
    \\[1.5em]

    \infer[]
    {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
    \Downarrow (\Theta' \mid v) : \tau}
    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta}\sigma =
    e}{e'})\Downarrow (\Theta' \mid v) : \tau
    }\\[1.5em]

    \infer[]
    {\,}
    {\Gamma ; \Delta \vdash (\Theta \mid K~x_1~\dots~x_n) \Downarrow
    (\Theta \mid K~x_1~\dots~x_n) : T}
    \\[1.5em]

    \infer[]
    {\Gamma;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' :
    K~x_1~\dots~x_n) : T \quad
    \Gamma ;\Delta \vdash (\Theta' \mid e'[x_i/y_i][K~x_1~\dots~x_n/z]
    ) \Downarrow (\Theta'' : v) :\tau }
    {\Gamma ;\Delta \vdash (\Theta \mid
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}})
    \Downarrow (\Theta'' \mid  v) : \tau}\\[1.5em]

    \infer[]
    {\Gamma ;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
    K~x_1~\dots~x_n) : T \quad
    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~x_1~\dots~x_n/z])
    \Downarrow (\Theta'' \mid v) : \tau }
    {\Gamma ; \Delta \vdash (\Theta \mid 
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
    \\[1.5em]
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
    e_i} \mid e' )\Downarrow (\Theta'
    \mid v) : \tau
    }
    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
    \Downarrow (\Theta' \mid v) : \tau}

   \end{array}
\]
